Transcriptomic responses to diet quality and viral infection in Apis mellifera

### 2018-10-26 Manuscript/Text/manuscript.tex
Added bigPint explanations

There are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees, but there have also been several such studies in model organisms. For example, diet effects on transcriptomics have been studied in insect models such as Drosophila. For instance, one recent transcriptomic study in Drosophila melanogaster reported an overexpression of genes related to immunity, metabolism, and hemocyanin in a high-fat diet and overexpression of genes related to cell cycle activity, DNA binding and transcription, and CHK kinase-like protein activity in a high-sugar diet \cite{sugarFat}. CHK kinase-like protein is believed to respond to DNA damage by arresting the cell cycle. This same study also discovered an upregulation of genes related to peptide and carbohydrate processing in both high-fat and high-sugar diets, a finding the authors attributed to a general increase in caloric intake. Another recent Drosophila study investigated the transcriptomic effects of diets high in protein relative to sugar, diets high in sugar relative to protein, and diets with equal amounts of protein and sugar \cite{sugarProtein}. No differential expression was found in Drosophila melanogaster, a species known to thrive in all three dietary conditions. However, Drosophila mojavensisa and Drosophila arizonae showed substantial differential expression between the dietary conditions: Genes involved in carbohydrate and lipid metabolism were upregulated in response to high sugar low protein diets and genes involved in juvenile hormone (JH) and ecdysone were upregulated in reponse to low sugar high protein diets. Interestingly, prior studies have suggested that JH regulates body size by controlling ecdysone production, which modifies insulin signaling \cite{jHormone}. One transcriptomic study in honey bees found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}.

Likewise, studies on the transcriptomic effects of virus infection have been performed in insect models such as Drosophila and mosquito. These studies have shown that RNA silencing is a major antiviral strategy, but other pathways play substantial roles in many virus-host systems like the Toll pathway, IMD pathway, JAK/STAT pathway, Toll-7-autophagy pathway, and transcriptional pausing \cite{cherryReview, swevers}. It is important to note that general immune responses to viral infection in insects might be an indirect result of cellular damage \cite{swevers}. In fact, every virus-host interaction has its own particularities derived of the diverse ways of replication and infection cycle evolved by different viruses. In this sense, an intricate set of pro- and anti-virus host factors are involved in each case \cite{cherryReview, swevers}. For example, dicistroviruses Drosophila C virus (DCV) and cricket paralysis virus (CrPV) infection requires ribosomal proteins RpS6, RpL19 and RACK1, but such proteins are dispensable during infection by the rhabdovirus vesicular stomatitis virus (VSV) in Drosophila. In contrast, the autophagy pathway has proven to be an effective antiviral strategy against VSV, but has no role limiting DCV or other positive-sense RNA viruses \cite{swevers}. For an extensive up-to-date review of known antiviral and proviral factors for specific viruses in Drosophila see \cite{swevers}. Model insect studies have shed light on how honey bees respond to viruses because the function of many genes relies on the characterization of orthologue genes. Nonetheless, recent transcriptomic studies in honey bees have improved our understanding of compatible and incompatible virus-bee interactions. Key factors or hallmarks of the antiviral conserved defense pathways mentioned above have been found in some virus-bee systems (e.g. RNA silencing, Toll, IMD, JAK/STAT pathway, autophagy, and endocytosis) \cite{McMenamin}. Added to these, a non-sequence-specific antiviral response mediated by unspecific dsRNA pathway was discovered in honey bees \cite{Flenniken}. This path has been studied recently in more depth using transcriptomics, which found known immune factors but also new ones for which indispensability in antiviral response was determined by silencing \cite{Brutscher}. Similar to Drosophila examples, particular antiviral factors in honey bees are linked to specific viruses, specific developmental stages, the analyzed tissue, the route of inoculation, and the time (post-inoculation) during which the study was performed. This was clearly demonstrated when comparing honey bee responses to two related iflaviruses with very different infection dynamics, sacbrood bee virus (SBV) vs. deformed wing virus (DWV) \cite{Ryabov}. Authors observed differences in induction of defensin and hymenoptaecin immune-related genes, and suggested the results reflect adaptations to the different routes of transmission \cite{Ryabov}. In the case of dicistroviruses, few works have studied the impact of IAPV infection at transcriptional level. Chen et al. 2014 analyzed responses to IAPV infection in larvae and workers using microarrays \cite{Chen}. Many of the DEGs found were involved in immune response and energy-related metabolism, particularly in adults but not in brood. The authors propose this observed difference could be connected to latent infections in larvae (where host immunity is not perturbed) versus acute infections in adulthood (induced by stressors faced during development) \cite{Chen}. IAPV acute infection also alters the DNA methylation pattern of numerous genes that do not overlap the genes that are up- or down-regulated at the transcriptional level \cite{galbraith}. These works reiterate the conclusion that viruses trigger particular antiviral mechanisms by different means and depending on several factors. The honey bee antiviral pathways induced by specific viruses were recently reviewed \cite{McMenamin}; it is noteworthy to mention that many honey bee factors discovered by transcriptomics need further characterization to uncover their role in controlling (or promoting) viral infection in honey bees.

@@

We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset. Note that the Galbraith dataset contained three samples for each virus level, while our dataset contained twelve samples for each virus level. Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may separate the infected and uninfected honey bees better than our dataset (Additional file 2). We also noted that the first replicate of both treatment groups in the Galbraith data did not cluster as cleanly in the PCA plots. However, through this automatically-generated plot, we can only visualize information at the sample level. Wanting to learn more about the data at the gene level, we continued with additional visualization techniques.

We used parallel coordinate lines superimposed onto boxplots to visualize the DEGs associated with virus infection in the two studies. The background side-by-side boxplot represents the distribution of all genes in the data, and each parallel coordinate line represents one DEG. To reduce overplotting of parallel coordinate lines, we used hierarchical clustering techniques to separate DEGs into common patterns as is described in the methods section.

We see that the 1,019 DEGs from the Galbraith dataset form relatively clean-looking visual displays (Figure 3). We do see that the first replicate of the virus group (V.1) appears somewhat inconsistent with the other virus replicates in Cluster 1, confirming that this trend in the data that we saw in the PCA plot carried through into the DEG results. In contrast, we see that the 43 virus-related DEGs from our dataset do not look as clean in their visual displays (Figure 4). The replicates appear somewhat inconsistent in their estimated expression levels and there is not always such a large difference between treatment groups. We see a similar finding when we also examine a larger subset of 1,914 diet-related DEGs from our study (Additional file 3).

We also used litre plots to examine the structure of individual DEGs: We see that indeed the individual virus DEGs from our data (Additional file 4) show less consistent replications and less differences between the treatment groups compared to the individual virus DEGs from the Galbraith data (Additional files 5 and 6). For the Galbraith data, we examined individual DEGs from the first cluster (Additional file 5) and second cluster (Additional file 6) because the first cluster had previously shown less consistency in the first replicate of the treatment group (Figure 3). We verify this trend again in the litre plots with the DEG points in the first cluster showing less tight cluster patterns (Additional files 5 and 6).

Finally, we looked at scatterplot matrices to assess the DEGs. We created standardized scatterplot matrices for each of the four clusters (from Figure 3) of the Galbraith data (Additional files 7, 8, 9, and 10). We also created standardized scatterplot matrices for our data. However, as our dataset contained 24 samples, we would need to include 276 scatterplots in our matrix, which would be too numerous to allow for efficient visual assessment of the data. As a result, we created four scatterplot matrices of our data, each with subsets of 6 samples to be more comparable to the Galbraith data (Additional files 11, 12, 13, and 14). We can again confirm through these plots that the DEGs from the Galbraith data appeared more as expected: They deviated more from the \textit{x=y} line in the treatment scatterplots while staying close to the \textit{x=y} line in replicate scatterplots.

>>>>

There are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees, but there have also been several such studies in model organisms. For example, diet effects on transcriptomics have been studied in the insect model Drosophila. One recent transcriptomic study in Drosophila melanogaster reported an overexpression of genes related to immunity, metabolism, and hemocyanin in a high-fat diet and overexpression of genes related to cell cycle activity, DNA binding and transcription, and CHK kinase-like protein activity in a high-sugar diet \cite{sugarFat}. CHK kinase-like protein is believed to respond to DNA damage by arresting the cell cycle. This same study also discovered an upregulation of genes related to peptide and carbohydrate processing in both high-fat and high-sugar diets, a finding the authors attributed to a general increase in caloric intake. Another recent Drosophila study investigated the transcriptomic effects of diets high in protein relative to sugar, diets high in sugar relative to protein, and diets with equal amounts of protein and sugar \cite{sugarProtein}. No differential expression was found in Drosophila melanogaster, a species known to thrive in all three dietary conditions. However, Drosophila mojavensisa and Drosophila arizonae showed substantial differential expression between the dietary conditions: Genes involved in carbohydrate and lipid metabolism were upregulated in response to high sugar low protein diets and genes involved in juvenile hormone (JH) and ecdysone were upregulated in reponse to low sugar high protein diets. Interestingly, prior studies have suggested that JH regulates body size by controlling ecdysone production, which modifies insulin signaling \cite{jHormone}. While similar transcriptomic studies have been limited in honey bees, one study found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}.

Likewise, studies on the transcriptomic effects of virus infection have been performed in insect models such as Drosophila and mosquito. These studies have shown that RNA silencing is a major antiviral strategy, but other pathways play substantial roles in many virus-host systems like the Toll pathway, IMD pathway, JAK/STAT pathway, Toll-7-autophagy pathway, and transcriptional pausing \cite{cherryReview, swevers}. It is important to note that general immune responses to viral infection in insects might be an indirect result of cellular damage \cite{swevers}. In fact, every virus-host interaction has its own particularities derived of the diverse ways of replication and infection cycle evolved by different viruses. In this sense, an intricate set of pro- and anti-virus host factors are involved in each case \cite{cherryReview, swevers}. For example, dicistroviruses Drosophila C virus (DCV) and cricket paralysis virus (CrPV) infection requires ribosomal proteins RpS6, RpL19 and RACK1, but such proteins are dispensable during infection by the rhabdovirus vesicular stomatitis virus (VSV) in Drosophila. In contrast, the autophagy pathway has proven to be an effective antiviral strategy against VSV, but has no role limiting DCV or other positive-sense RNA viruses \cite{swevers}. For an extensive up-to-date review of known antiviral and proviral factors for specific viruses in Drosophila see \cite{swevers}. Model insect studies have shed light on how honey bees respond to viruses because the function of many genes relies on the characterization of orthologue genes. Nonetheless, recent transcriptomic studies in honey bees have improved our understanding of compatible and incompatible virus-bee interactions. Key factors or hallmarks of the antiviral conserved defense pathways mentioned above have been found in some virus-bee systems (e.g. RNA silencing, Toll, IMD, JAK/STAT pathway, autophagy, and endocytosis) \cite{McMenamin}. Added to these, a non-sequence-specific antiviral response mediated by unspecific dsRNA pathway was discovered in honey bees \cite{Flenniken}. This path has been studied recently in more depth using transcriptomics, which found known immune factors but also new ones for which indispensability in antiviral response was determined by silencing \cite{Brutscher}. Similar to Drosophila examples, particular antiviral factors in honey bees are linked to specific viruses, specific developmental stages, the analyzed tissue, the route of inoculation, and the time (post-inoculation) during which the study was performed. This was clearly demonstrated when comparing honey bee responses to two related iflaviruses with very different infection dynamics, sacbrood bee virus (SBV) vs. deformed wing virus (DWV) \cite{Ryabov}. Authors observed differences in induction of defensin and hymenoptaecin immune-related genes, and suggested the results reflect adaptations to the different routes of transmission \cite{Ryabov}. In the case of dicistroviruses, few works have studied the impact of IAPV infection at transcriptional level. Chen et al. 2014 analyzed responses to IAPV infection in larvae and workers using microarrays \cite{Chen}. Many of the DEGs found were involved in immune response and energy-related metabolism, particularly in adults but not in brood. The authors propose this observed difference could be connected to latent infections in larvae (where host immunity is not perturbed) versus acute infections in adulthood (induced by stressors faced during development) \cite{Chen}. IAPV acute infection also alters the DNA methylation pattern of numerous genes that do not overlap the genes that are up- or down-regulated at the transcriptional level \cite{galbraith}. These works reiterate the conclusion that viruses trigger particular antiviral mechanisms by different means and depending on several factors. The honey bee antiviral pathways induced by specific viruses were recently reviewed \cite{McMenamin}; it is noteworthy that many honey bee factors discovered by transcriptomics need further characterization to uncover their role in controlling (or promoting) viral infection in honey bees.

@@

We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset. Note that the Galbraith dataset contained three samples for each virus level, while our dataset contained twelve samples for each virus level. Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may separate the infected and uninfected honey bees better than our dataset (Additional file 2). We also noted that the first replicate of both treatment groups in the Galbraith data did not cluster as cleanly in the PCA plots. However, through this automatically-generated plot, we can only visualize information at the sample level. Wanting to learn more about the data at the gene level, we continued with new visualization techniques that we recently published in our R package bigPint (https://lrutter.github.io/bigPint).

We used parallel coordinate lines superimposed onto side-by-side boxplots to visualize the DEGs associated with virus infection in the two studies. The background side-by-side boxplot represents the distribution of \textit{all} genes in the data, and each parallel coordinate line represents one DEG. In a parallel coordinate line, connections between samples with positive correlations should be flat, while connections between samples with negative correlations should be crossed. We expect DEGs to show more variability between treatments than between replicates. This means the parallel coordinate lines should be flat between replicates but crossed between treatments. However, overplotting problems would obscure our visualization if we were to plot all DEGs onto the same side-by-side boxplot. As a result, we used hierarchical clustering techniques to separate DEGs into common patterns as is described in the methods section.

We see that the 1,019 DEGs from the Galbraith dataset form relatively clean-looking visual displays, with consistent replicates and differences between treatments (Figure 3). We do see that the first replicate of the virus group (V.1) appears somewhat inconsistent with the other virus replicates in Cluster 1, confirming that the trend we saw in the PCA plot carried through into the DEG results. Cluster 4 reveals somewhat inconsistent replicates in the virus group, although most virus standardized read counts (group V) remain consistently larger than most control standardized read counts (group N). In contrast, we see that the 43 virus-related DEGs from our dataset do not look as clean in their visual displays (Figure 4). The replicates appear somewhat inconsistent in their estimated expression levels and there is not always such a large (or even consistent) difference between treatment groups. We see a similar finding when we also examine a larger subset of 1,914 diet-related DEGs from our study (Additional file 3).

We next used repLIcate TREatment (``litre'') plots, which we recently developed and published in our bigPint software package. Litre plots allow users to visualize one DEG onto the Cartesian coordinates of one scatterplot matrix. In the litre plot, each gene in the data is plotted once for every combination of replicates between treatment groups. For example, there are nine ways to pair a replicate from one treatment group with a replicate from the other treatment group in the Galbraith dataset (N.1 and V.1, N.1 and V.2, N.1 and V.3, N.2 and V.1, N.2. and V.2, N.2 and V.3, N.3 and V.1, N.3 and V.2, and N.3 and V.3). Hence, each gene in the Galbraith dataset is plotted as nine points in the litre plot. With 11,825 genes in the Galbriath data, 106,425 points would need to be plotted. Our dataset is even more dramatic: There are 144 ways to pair a replicate from one treatment group with a replicate from the other treatment group, and with 15,314 genes in our data, we would need to plot 2,205,216 points. For either dataset, plotting all these points would reduce the speed of interactive functionality and cause overplotting problems. As a result, we use hexagon bins to summarize this massive information. Once the background of hexagons has been drawn to reveal the distribution of all between-treatment sample pair combinations for \textit{all} genes, the user can superimpose all between-treatment sample pair combinations for one gene of interest.

Additional file 4 shows nine example litre plots for our dataset. The hexagon background is the same for all nine litre plots because it simply shows the distribution of all between-treatment sample pair combinations for \textit{all} genes in our dataset. In each litre plot, there are 144 magenta points superimposed that show all between-treatment sample pair combinations for one DEG of interest. Additional file 5 and 6 similarly each show nine example litre plots for the Galbraith datset. We examined individual DEGs from the first cluster (Additional file 5) and second cluster (Additional file 6) of the Galbriath data because the first cluster had previously shown less consistency in the first replicate of the treatment group (Figure 3). Notice that, as previously explained, we now show each DEG as nine points for the Galbraith dataset. We see that indeed the virus DEGs from our data (Additional file 4) show less consistent replications and less differences between the treatment groups compared to the virus DEGs from the Galbraith data (Additional files 5 and 6). We also observe that, in the Galbraith dataset, the DEG points in the first cluster show less tight cluster patterns than the DEG points in the second cluster (Additional files 5 and 6), an observation we saw previously in the parallel coordinate plots (Figure 3).

Finally, we used scatterplot matrices from the bigPint software to further assess the DEGs. A scatterplot matrix is another effective multivariate visualization tool that plots read count distributions across all genes and samples. Specifically, it represents every gene in the dataset as a black point in each scatterplot. DEGs can be superimposed as colored points to assess their patterns against the full dataset. We expect DEGs to mostly fall along the \textit{x=y} line in replicate scatterplots (denoting replicate consistency) but deviate from the \textit{x=y} line in treatment scatterplots (denoting significant treatment changes). The \textit{x=y} line is shown in red in our plots.

We created standardized scatterplot matrices for each of the four clusters (from Figure 3) of the Galbraith data (Additional files 7, 8, 9, and 10). We also created standardized scatterplot matrices for our data. However, as our dataset contained 24 samples, we would need to include 276 scatterplots in our matrix, which would be too numerous to allow for efficient visual assessment of the data. As a result, we created four scatterplot matrices of our data, each with subsets of 6 samples to be more comparable to the Galbraith data (Additional files 11, 12, 13, and 14). We can again confirm through these plots that the DEGs from the Galbraith data appeared more as expected: They deviated more from the \textit{x=y} line in the treatment scatterplots while staying close to the \textit{x=y} line in replicate scatterplots.
